---
title: "Excess deaths 2021"
output: html_notebook
---


```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(janitor)
library(lubridate)
library(stringr)
library(kableExtra)
library(formattable)
library(DatawRappr)
library(ggthemes)
library(RMySQL)
library(forcats)
library(MMWRweek)
library(scales)
library(readxl)


subtitle <-  '52 weeks of year'
wknumber <- 52


mnpop_by_age <- read_xlsx('./data/MNpop_SingleYearofAge_1980_2019.xlsx', sheet="data") %>% clean_names()


mn_pop_groups <- mnpop_by_age %>% group_by(year, age_group) %>% summarise(tot = sum(estimate)) %>% mutate(yr = as.character(year))



# create age groups for 2020 and 2021

```



```{r}

#the 2021 death data was imported to the mySQL server
#I combined numerous years of data into a table called "deaths_illdefined_analysis" to use for this
#and another story that I might do
#this chunk of code grabs a few years of that data, and only certain columns


con <- dbConnect(RMySQL::MySQL(), host = Sys.getenv("host"), dbname="newsroomdata",user= Sys.getenv("userid"), password=Sys.getenv("pwd"))

#list the tables in the database we've connected to
#dbListTables(con)

#list the fields in the table; change "mytablename" to the name of the table you're trying to connect to
#dbListFields(con,'mytablename')


#Pull selected data using query inside the parentheses
data1 <- dbSendQuery(con, "select  ST_FILE_NBR,DECD_FRST_NME,DECD_MIDD_NME,DECD_LST_NME,DECD_MAIDN_NME,DECD_SUFX,DECD_AGE_YR,
DECD_BRTH_DT,DECD_DTH_DT,DECD_BRTH_CNTRY,DECD_RES_NURSING,DECD_RES_FCLTY,DECD_RES_CTY,
DECD_RES_ST,DECD_RES_ZIP5,DECD_RES_CNTY,DECD_RES_CNTY_FIPS_CD,DECD_MRTL_STATUS,
CAUSE_DTH_A,CAUSE_INTRVL_A,CAUSE_DTH_B,CAUSE_INTRVL_B,CAUSE_DTH_C,CAUSE_INTRVL_C,CAUSE_DTH_D,CAUSE_INTRVL_D,CAUSE_DTH_OTHR,
CERTFR_MANNER_DTH,DT_FILED_MED,DT_FILED_LEGL,CMPLT_FL,
SUBJECT_NOT_HISPANIC,SUBJECT_MEXICAN,SUBJECT_PUERTO_RICAN,SUBJECT_CUBAN,SUBJECT_OTHER_SPANISH,SUBJECT_OTHR_SPAN_TXT,
SUBJECT_REFUSED_HISPANIC,SUBJECT_UNKNOWN_HISPANIC,SUBJECT_NOTOBTAINABLE_HISPANIC,SUBJECT_WHITE,SUBJECT_AFRICAN_AMERICAN,SUBJECT_SOMALI,SUBJECT_ETHIOPIAN,
SUBJECT_LIBERIAN,SUBJECT_KENYAN,SUBJECT_SUDANESE,SUBJECT_NIGERIAN,SUBJECT_GHANIAN,SUBJECT_OTHER_AFRICAN,SUBJECT_OTHER_AFRICAN_TXT,SUBJECT_AMERICAN_INDIAN,
SUBJECT_TRIBE_TXT1,SUBJECT_TRIBE_TXT2,SUBJECT_ASIAN_INDIAN,SUBJECT_CHINESE,SUBJECT_FILIPINO,SUBJECT_JAPANESE,SUBJECT_KOREAN,SUBJECT_VIETNAMESE,
SUBJECT_HMONG,SUBJECT_CAMBODIAN,SUBJECT_LAOTIAN,SUBJECT_OTHER_ASIAN,SUBJECT_OTHR_ASN_TXT1,SUBJECT_OTHR_ASN_TXT2,SUBJECT_HAWAIIAN,SUBJECT_GUAMANIAN_CHAMORRO,
SUBJECT_SAMOAN,SUBJECT_OTHER_PACIFIC_ISLANDER,SUBJECT_OTHR_PAC_ISLE_TXT1,SUBJECT_OTHR_PAC_ISLE_TXT2,SUBJECT_OTHER,SUBJECT_OTHR_TXT1,
SUBJECT_OTHR_TXT2,SUBJECT_REFUSED_RACE,SUBJECT_UNKNOWN,SUBJECT_NOTOBTAINABLE_RACE,SUBJECT_RACE_MVR,
RAC_ICD01,RAC_CREATED01,RAC_ICD02,
RAC_CREATED02,RAC_ICD03,RAC_CREATED03,RAC_ICD04,RAC_CREATED04,RAC_ICD05,RAC_CREATED05,RAC_ICD06,RAC_CREATED06,RAC_ICD07,RAC_CREATED07,RAC_ICD08,
RAC_CREATED08,RAC_ICD09,RAC_CREATED09,RAC_ICD10,RAC_CREATED10,ACME_UNLY_CAUSE_DTH,MNL_UNLY_CAUSE_DTH,RACE,HISPANICETHNICITY,YR, covid2, mdhgroup, unly_cause_desc, opioids
from deaths_illdefined_analysis
where yr=2017 or yr=2018 or yr=2019 or yr=2020 or yr=2021")

#assign it to a new data frame
deaths <- fetch(data1, n=-1)

dbClearResult(data1)


#disconnect connection
dbDisconnect(con)



#####DATA CLEANUP:


deaths <- deaths %>% clean_names() %>%  mutate(decd_dth_dt = lubridate::as_date(decd_dth_dt),
                            decd_brth_dt = lubridate::as_date(decd_brth_dt),
                            dt_filed_legl = lubridate::as_date(dt_filed_legl),
                            dt_filed_med = lubridate::as_date(dt_filed_med)) 


deaths$certfr_manner_dth[is.na(deaths$certfr_manner_dth)] <- 'not listed'








#The first day of any MMWR week is Sunday. MMWR week numbering is sequential beginning with 1 and
#incrementing with each week to a maximum of 52 or 53. MMWR week #1 of an MMWR year is the first week of
#the year that has at least four days in the calendar year. For example, if January 1 occurs on a Sunday, Monday,
#Tuesday or Wednesday, the calendar week that includes January 1 would be MMWR week #1. If January 1
#occurs on a Thursday, Friday, or Saturday, the calendar week that includes January 1 would be the last MMWR
#week of the previous year (#52 or #53). Because of this rule, December 29, 30, and 31 could potentially fall into
#MMWR week #1 of the following MMWR year. 

#this uses the MMWRweek package to assign the month, week and day from each date
# the MMWRweek package creates a whole new dataframe when it does this so you first need to do that

mmwrdates <- MMWRweek(deaths$decd_dth_dt)

#then join (or bind) those columns to the covid data 
#this also uses the clean_names function from janitor to lowercase the new columns
deaths <- cbind(deaths, mmwrdates) %>% clean_names()


#create a table that has the start and end dates of each week
#this will be used again later
week_start_end <-  deaths %>%
  group_by(mmw_ryear, mmw_rweek) %>%
  summarise(start_date = min(decd_dth_dt),
            end_date = max(decd_dth_dt),
            .groups='drop')


# apply the start_date and end_date to covid table
 deaths <-  left_join(deaths, week_start_end %>% select(mmw_ryear, mmw_rweek, start_date, end_date), by=c("mmw_ryear"="mmw_ryear", "mmw_rweek"="mmw_rweek" ))




#categorize deaths by COVID or the manner of death for all other types of death 
 deaths <-  deaths %>% mutate(death_type = case_when(covid2=='covid-unly'~ 'COVID',
                                                     covid2!='covid-unly' | is.na(covid2) & certfr_manner_dth=='ACCIDENT' ~ 'Accident',
                                                     covid2!='covid-unly'| is.na(covid2) & certfr_manner_dth=='SUICIDE' ~ 'Suicide',
                                                     covid2!='covid-unly' | is.na(covid2)& certfr_manner_dth=='NATURAL' ~ 'Natural',
                                                     covid2!='covid-unly' | is.na(covid2)& certfr_manner_dth=='HOMICIDE' ~ 'Homicide',
                                                     covid2!='covid-unly' | is.na(covid2) & certfr_manner_dth %in% c("COULD NOT BE DETERMINED", "PENDING INVESTIGATION") ~ 'Accident',
                                                     TRUE ~ 'unknown'))
 
 
 #this is a more refined categorization where it breaks out the COVID deaths into those
 #that more expressly define COVID deaths where it was listed as the underlying cause of death and others that didn't
 
  deaths <-  deaths %>% mutate(death_type2 = case_when(covid2 %in% c('cov-no cod', 'covid-othr', 'presumed')~ 'COVID-other',
                                                       covid2=='covid-unly' ~ 'COVID-underlying',
                                                     covid2!='covid-unly'| is.na(covid2) & certfr_manner_dth=='ACCIDENT' ~ 'Accident',
                                                     covid2!='covid-unly'| is.na(covid2) & certfr_manner_dth=='SUICIDE' ~ 'Suicide',
                                                     covid2!='covid-unly' | is.na(covid2)& certfr_manner_dth=='NATURAL' ~ 'Natural',
                                                     covid2!='covid-unly' | is.na(covid2)& certfr_manner_dth=='HOMICIDE' ~ 'Homicide',
                                                     covid2!='covid-unly' | is.na(covid2) & certfr_manner_dth %in% c("COULD NOT BE DETERMINED", "PENDING INVESTIGATION") ~ 'Accident',
                                                     TRUE ~ 'unknown'))


 

```

# Total deaths by year
```{r}
deaths %>% group_by(yr) %>% summarise(count=n())
```



# Deaths by week
```{r}

byweek2021 <-  deaths %>% filter(mmw_rweek<=wknumber, mmw_ryear==2021) %>% group_by(mmw_rweek, death_type) %>% summarise(tot=n())

byweek <-  deaths %>% filter(mmw_rweek<=wknumber) %>% group_by(mmw_ryear, mmw_rweek) %>% summarise(tot=n())

byweekavg <-  byweek %>% filter(mmw_rweek<=wknumber, mmw_ryear>=2010 & mmw_ryear<2020) %>% group_by(mmw_rweek) %>% summarise(avg = mean(tot))

byweek2020 <- deaths %>% filter(mmw_rweek<=wknumber, mmw_ryear==2020) %>% group_by(mmw_ryear) %>% summarise(tot=n())


byweek_lines <- deaths %>%  filter(mmw_rweek<=wknumber, mmw_ryear %in% c("2020", "2021")) %>% group_by(mmw_ryear, mmw_rweek) %>% summarise(tot=n()) %>% mutate(mmw_ryear_txt=as.character(mmw_ryear))

byweekavg <-  byweekavg %>% mutate(mmw_ryear_txt='Avg 2010-2019', tot=avg)

byweek_lines <- bind_rows(byweek_lines, byweekavg %>% select(-avg)) 


ggplot(byweek_lines, aes(x=byweek_lines$mmw_rweek, y=tot, group=mmw_ryear_txt, color=mmw_ryear_txt))+
  geom_line(stat="identity", size=1)+
  scale_y_continuous(name="Total deaths", limits=c(0, 1500), breaks=seq(0, 1500, 300))+
  scale_x_continuous(name="Week of the year", breaks=seq(1, wknumber, 2))+
     theme(legend.position="bottom",legend.title = element_blank())+
    theme_hc()+
      labs(title = "Deaths by week", 
       subtitle = subtitle,
       caption = "Source: Minnesota death certificates")
    

  
```


```{r}
ggsave("./output/total_weekly_deaths_2021.jpg", width=8, height=5, units="in", dpi="print")

write.csv(pivot_wider(byweek_lines %>% ungroup() %>% select(-mmw_ryear), names_from ="mmw_ryear_txt", values_from="tot"), './output/table_deaths_by_week_2021.csv', row.names=FALSE)
```


# Deaths by type
```{r}

write.csv(pivot_wider(deaths %>% filter(yr>=2017) %>% 
  group_by(yr, death_type) %>% summarise(count=n()), names_from=yr, values_from=count), './output/death_types_2021.csv', row.names=FALSE)

pivot_wider(deaths %>% filter(yr>=2017) %>% 
  group_by(yr, death_type) %>% summarise(count=n()), names_from=yr, values_from=count)


```

```{r}
write.csv(pivot_wider(deaths %>% filter(yr>=2017) %>% 
  group_by(yr, death_type2) %>% summarise(count=n()), names_from=yr, values_from=count), './output/alternate_death_types_2021.csv', row.names=FALSE)



pivot_wider(deaths %>% filter(yr>=2017) %>% 
  group_by(yr, death_type2) %>% summarise(count=n()), names_from=yr, values_from=count)
```

```{r}
deaths$death_type2 <- factor(deaths$death_type2, levels=c("COVID-other",  "COVID-underlying", "unknown", "Homicide","Suicide", "Accident","Natural"  ))


ggplot(deaths %>% filter(yr>=2017) %>% group_by(yr, death_type2) %>% summarise(count=n()) , aes(x=yr, y=count, fill=death_type2))+
    geom_col(stat="identity", position="stack")+
    scale_fill_manual(values = c("#ef8a62", "#b2182b", "#999999", "#1b7837","#762a83", "#67a9cf", "#f6e8c3" ))+
  scale_y_continuous(limits=c(0,55000), breaks=seq(0,55000, 5000), labels=scales::comma)+
  #theme_hc()+
  labs(title="Minnesota deaths by type",
        subtitle="2017-2021")


ggsave("./output/total_deaths_by_type_2021.jpg", width=8, height=5, units="in", dpi="print")
```



# Accidental deaths

```{r}


deaths %>% filter(opioids=='true', death_type2=='Accident') %>% group_by(yr) %>% summarise(count=n())

```


# Opioid deaths
```{r}

opioid_deaths <-  deaths %>% filter(opioids=='true', yr!=2008) %>% group_by(yr) %>% summarise(num_deaths=n()) %>% mutate(yr=as.numeric(yr))


ggplot(data=opioid_deaths, aes(x=yr, y=num_deaths))+
         geom_line(stat="identity")+
  scale_y_continuous(limits=c(0,1200), breaks=seq(0,1200,100), name="Number of deaths")+
  scale_x_continuous(breaks=seq(2009,2021, 1), name="Year")+
  labs(title="Opioid-related deaths in MN",
       subtitle="2008-2021",
       caption= "2021 is preliminary")+
  theme_hc()

ggsave("./output/opioid_deaths_2008_2021.jpg", width=8, height=5, units="in", dpi="print")
```

# COVID deaths - average age over time

```{r}

                                                   
                                                   
ggplot(deaths %>% filter(covid2=='covid-unly') %>% mutate(mnthyr =  floor_date(decd_dth_dt, "month")) %>% group_by(mnthyr) %>% summarise(avgage = mean(decd_age_yr)), aes(x=mnthyr, y=avgage))+
  geom_line(stat="identity")+
  scale_y_continuous(limits=c(0,100))+
      scale_x_date(date_breaks = "3 months", date_labels="%m-%y")+
  labs(title="COVID deaths average age",
       subtitle = "COVID as underlying cause of death" )+
  theme_hc()

```


# COVID deaths under age 65

```{r}
ggplot(deaths %>% filter(covid2=='covid-unly', decd_age_yr<65) %>% mutate(mnthyr =  floor_date(decd_dth_dt, "month")) %>% group_by(mnthyr) %>% summarise(num_deaths = n()), aes(x=mnthyr, y=num_deaths))+
  geom_bar(stat="identity")+
  scale_y_continuous(limits=c(0,200))+
      scale_x_date(date_breaks = "3 months", date_labels="%m-%y")+
  labs(title="COVID deaths under age 65",
       subtitle = "COVID as underlying cause of death" )+
  theme_hc()
```


```{r}
ggplot(deaths %>% filter(covid2=='covid-unly', decd_age_yr<65) %>%  group_by(yr) %>% summarise(num_deaths = n()), aes(x=yr, y=num_deaths))+
  geom_bar(stat="identity")+
  scale_y_continuous(limits=c(0,1200))+
      #scale_x_date(date_breaks = "3 months", date_labels="%m-%y")+
  labs(title="COVID deaths under age 65",
       subtitle = "COVID as underlying cause of death" )+
  theme_hc()
```

# Deaths by age groups

```{r}

deaths <-  deaths %>% mutate(age_group = case_when(decd_age_yr<18 ~ 'Under 18',
                                                   decd_age_yr>=18 & decd_age_yr<=24 ~ '18 to 24',
                                                   decd_age_yr>=25 & decd_age_yr<35 ~ '25 to 34',
                                                   decd_age_yr>=35 & decd_age_yr<45 ~ '35 to 44',
                                                   decd_age_yr>=45 & decd_age_yr<55 ~ '45 to 54',
                                                   decd_age_yr>=55 & decd_age_yr<65 ~ '55 to 64',
                                                   decd_age_yr>=65 & decd_age_yr<75 ~ '65 to 74',
                                                   decd_age_yr>=75 & decd_age_yr<85 ~ '75 to 84',
                                                   decd_age_yr>=85  ~ '85 and up',
                                                   TRUE ~ 'unknown'))





by_agegroup <-  deaths %>% filter(yr>=2018, decd_age_yr<65) %>%  group_by(yr, age_group) %>% summarise(total = n())

by_agegroup <-  left_join(by_agegroup, mn_pop_groups, by=c("yr"="yr", "age_group"="age_group"))
write.csv(deaths %>% group_by(yr, age_group) %>% summarise(total=n()), './output/by_agegroup_2021.csv', row.names=FALSE)

ggplot(by_agegroup, aes(x=by_agegroup$yr, y=total, group=age_group, color=age_group))+
  geom_line(stat="identity", size=1)+
  #scale_y_continuous(name="Total deaths", limits=c(0, 1500), breaks=seq(0, 1500, 300))+
  #scale_x_continuous(name="Week of the year", breaks=seq(1, wknumber, 2))+
     theme(legend.position="bottom",legend.title = element_blank())+
    theme_hc()+
      labs(title = "Deaths by age group", 
       subtitle = subtitle,
       caption = "Source: Minnesota death certificates")


#pivot_wider(deaths %>% filter(opioids=='true', yr %in% c(2017, 2018, 2019)) %>% group_by(yr, age_group) %>% summarise(count=n()), names_from="yr", values_from="count") 

```



# Deaths among age 45 to 64
```{r}



deaths <-  deaths %>% mutate(death_big_type = case_when(death_type2 %in% c("COVID-other", "COVID-underlying") ~ 'COVID', 
                                                        TRUE ~ 'all other'))

deaths$death_big_type <- factor(deaths$death_big_type, levels=c("COVID", "all other" ))


ggplot(deaths %>% filter(age_group %in% c("55 to 64", "45 to 54")) %>%  group_by(yr, death_big_type) %>% summarise(num_deaths = n()), aes(x=yr, y=num_deaths, fill=death_big_type))+
  geom_bar(stat="identity", position="stack")+
 # scale_y_continuous(limits=c(0,1200))+
      #scale_x_date(date_breaks = "3 months", date_labels="%m-%y")+
  labs(title="Deaths age 45 to 64",
       subtitle = "All causes" )+
  theme_hc()




```

# Hispanic deaths

```{r}
deaths %>% filter(hispanicethnicity=='HISPANIC') %>% group_by(yr) %>% summarise(count=n())


ggplot(deaths %>% filter(hispanicethnicity=='HISPANIC') %>% group_by(yr) %>% summarise(count=n()), aes(x=yr, y=count))+
  geom_bar(stat="identity")+
  labs(title='Hispanic deaths by year')
```


```{r}

# in mySQL, I coded the deaths in a column called mdhgroup for the top 10 death categories (heart disease, cancer, accidents, etc). that column is stored in the ICD lookup table called icdcodes_revised_feb2022


mdhgroup_deaths <-  pivot_wider(deaths %>% group_by(yr, mdhgroup) %>% summarise(count=n()), names_from="yr", values_from="count") %>% clean_names()

#write.csv(mdhgroup_deaths, './output/mdhgroup_deaths_2021.csv', row.names=FALSE)
```


# Retrieve population by age data

```{r}
library(tidycensus)
#retrieve census api key from my Renviron file
census_api_key(Sys.getenv("CENSUS_API_KEY"))


#load the variable list from census
variable_list <- load_variables("2010", "sf1",  cache = FALSE)


#pull from the variable list the specific ones I want (from single-year-by age, with race breakdowns)
save_variables <-  variable_list %>% filter(str_sub(name, 1,6)== 'PCT012') %>% pull(name)

#pull the data using get_decennial function from tidycensus
#us_age_singleyr <-  get_decennial(geography = "us", variables = save_variables, year = "2010")
age_singleyr <-  get_decennial(geography = "state", state="MN", variables = save_variables, year = "2010")


age_singleyr2 <-  left_join(age_singleyr, variable_list, by=c("variable"="name")) %>%  #join with variables
  clean_names()%>%  #clean the headers
  separate("label", sep="!!", c("grp", "gender", "age"))%>%  #separate the label column
  filter(age!='NA', gender!='NA') %>%   #eliminate the total rows that we don't want to keep   
  mutate(age2 = case_when(age=='Under 1 year' ~ 0,
                          age %in% c('100 to 104 years', '105 to 109 years', '110 years and over') ~ 100,
                          age=='1 year'~ 1,
                          str_detect(age,"years") ~ as.numeric(str_replace(age, " years", ""))))


# create age by race/ethnicity table
race_age <-  age_singleyr2 %>%
  filter(str_sub(variable,1,7) %in% c("PCT012H", "PCT012I", "PCT012J","PCT012K","PCT012L","PCT012M","PCT012N","PCT012O" )) %>%
  mutate(race_group = str_sub(concept, 12,200),
         race_group = str_replace(race_group, '\\(', ''),
         race_group = str_replace(race_group, '\\)', ''),
         big_group = case_when(race_group=='WHITE ALONE, NOT HISPANIC OR LATINO' ~ 'WHITE',
                               TRUE ~ 'POC'),
         race_ethnicity = case_when(str_detect(race_group, "INDIAN")~ 'American Indian',
                                    str_detect(race_group, "ASIAN") ~'Asian',
                                    str_detect(race_group, 'BLACK')~ 'Black',
                                    race_group=='HISPANIC OR LATINO'~ 'Hispanic',
                                    str_detect(race_group, 'NATIVE HAWAIIAN')~ 'Pacific Islander',
                                    str_detect(race_group, 'SOME OTHER')~ 'Other',
                                    str_detect(race_group, 'TWO OR MORE')~ 'Multi',
                                    str_detect(race_group, 'WHITE ALONE')~'White'))


race_age <- race_age %>% mutate(agegroup = case_when(age2<=14 ~ '0-14',
                                                     age2>=15 & age2 <=19 ~ '15-19',
                                                     age2>=20 & age2<=24 ~ '20-24',
                                                     age2>=25 & age2<=29 ~ '25-29',
                                                     age2>=30 & age2<=34 ~ '30-34',
                                                     age2>=35 & age2<=39 ~ '35-39',
                                                     age2>=40 & age2<=44 ~ '40-44',
                                                     age2>=45 & age2<=49 ~ '45-49',
                                                     age2>=50 & age2<=54 ~ '50-54',
                                                     age2>=55 & age2<=59 ~ '55-59',
                                                     age2>=60 & age2<=64 ~ '60-64',
                                                     age2>=65 & age2<=69 ~ '65-69',
                                                     age2>=70 & age2<=74 ~ '70-74',
                                                     age2>=75 & age2<=79 ~ '75-79',
                                                     age2>=80 & age2<=84 ~ '80-84',
                                                     age2>=85 ~ '85+'))


race_age_totals <-  race_age %>% group_by(race_ethnicity, agegroup) %>% summarise(tot = sum(value))  

race_age_totals <-  race_age_totals %>% mutate(race_ethnicity=toupper(race_ethnicity))




```



# Age-adjusted death rates
```{r}

deaths <-   deaths  %>% 
  mutate(newrace = case_when(hispanicethnicity=='HISPANIC' ~ 'HISPANIC',
                             hispanicethnicity!='HISPANIC' & race=='Other Race' ~ 'OTHER',
                             hispanicethnicity!='HISPANIC' & race=='Other Asian' ~ 'ASIAN',
                           hispanicethnicity!='HISPANIC' ~ toupper(race),
                                      TRUE ~ 'unknown'),
         new_age_group = case_when(decd_age_yr<15 ~ '0-14',
                                              decd_age_yr>=15 & decd_age_yr<=19 ~ '15-19',
                                              decd_age_yr>=20 & decd_age_yr<=24 ~ '20-24',
                                   decd_age_yr>=25 & decd_age_yr<=29 ~ '25-29',
                                           decd_age_yr>=30 & decd_age_yr<=34 ~ '30-34',
                                    decd_age_yr>=35 & decd_age_yr<=39 ~ '35-39',
                                           decd_age_yr>=40 & decd_age_yr<=44 ~ '40-44',
                                    decd_age_yr>=45 & decd_age_yr<=49 ~ '45-49',
                                           decd_age_yr>=50 & decd_age_yr<=54 ~ '50-54',
                                    decd_age_yr>=55 & decd_age_yr<=59 ~ '55-59',
                                           decd_age_yr>=60 & decd_age_yr<=64 ~ '60-64',
                                    decd_age_yr>=65 & decd_age_yr<=69 ~ '65-69',
                                           decd_age_yr>=70 & decd_age_yr<=74 ~ '70-74',
                                    decd_age_yr>=75 & decd_age_yr<=79 ~ '75-79',
                                           decd_age_yr>=80 & decd_age_yr<=84 ~ '80-84',
                                               decd_age_yr>=85 ~ '85+',
                                           TRUE ~ 'unk'))




deaths_age_race_totals <-  deaths %>% group_by(yr, newrace, new_age_group) %>% summarise(deaths=n())


deaths_age_race_totals <-  deaths_age_race_totals %>% mutate(newrace=toupper(newrace))

deaths_age_race_totals <-  left_join(race_age_totals, deaths_age_race_totals,  by=c("race_ethnicity"="newrace", "agegroup"="new_age_group"))


standard_pop <-  read_csv('./data/us_standard_pop_weights.csv')

deaths_age_race_totals <-  left_join(deaths_age_race_totals, standard_pop, by=c("agegroup"="agegroup"))

deaths_age_race_totals <-  deaths_age_race_totals %>% mutate(crude_rate = (deaths/tot)*10000,
                                                             age_adjusted = crude_rate * weight)


ggplot(deaths_age_race_totals %>% group_by(yr, race_ethnicity) %>% summarise(adjusted = sum(age_adjusted)) %>% filter(yr!='NA', race_ethnicity!='OTHER', race_ethnicity!='PACIFIC ISLANDER'), aes(x=yr, y=adjusted, group=race_ethnicity, col=race_ethnicity))+
  geom_line(stat="identity")+
  labs(title="Age-adjusted death rates by group",
       subtitle="Per 10k")



#write.csv(deaths_age_race_totals, './output/deaths_age_race_totals.csv', row.names=FALSE)

#write.csv(race_age_totals, './output/race_age_population.csv', row.names=FALSE)
```




```{r}
#write.csv(deaths_age_race_totals %>% group_by(yr, race_ethnicity) %>% summarise(adjusted = sum(age_adjusted)) %>% filter(yr!='NA', race_ethnicity!='OTHER', race_ethnicity!='PACIFIC ISLANDER'), './output/age_adjusted_death_rates_by_group.csv', row.names=FALSE)
```

# Crude death rates

```{r}
deaths_age_race_totals  %>% group_by(yr, race_ethnicity) %>% summarise(pop = sum(tot),
                                                                                      dths = sum(deaths)) %>% 
  mutate(crude_rate = (dths/pop)*10000) %>% 
  arrange(race_ethnicity, yr)
```


```{r}
write.csv(deaths_age_race_totals  %>% group_by(yr, agegroup) %>% summarise(pop = sum(tot),
                                                                                      dths = sum(deaths)) %>% 
  mutate(crude_rate = (dths/pop)*10000) %>% 
  arrange(agegroup, yr), './output/crude_death_rates_by_age.csv', row.names=FALSE)
```


# COVID death rates by group
```{r}

covid_race_totals <-  deaths %>% filter(covid2 %in% c("covid-unly", "covid-othr")) %>%  group_by(newrace, new_age_group) %>% summarise(deaths=n())


covid_race_totals <-  covid_race_totals %>% mutate(newrace=toupper(newrace))

covid_race_totals <-  left_join(race_age_totals, covid_race_totals,  by=c("race_ethnicity"="newrace", "agegroup"="new_age_group"))

covid_race_totals$deaths[is.na(covid_race_totals$deaths)] <-  0

covid_race_totals %>% group_by(race_ethnicity) %>% summarise(pop = sum(tot), dths = sum(deaths)) %>% mutate(rate100k = (dths/pop)*100000 )
```




# Hispanic COVID deaths
```{r}
hispanic <-  deaths %>% filter(newrace=='HISPANIC')

hispanic %>% filter(covid2 %in% c('covid-unly', 'covid-othr')) %>% 
  group_by(yr) %>% summarise(count=n())
```

# COVID deaths age 50 and under
```{r}
deaths %>% filter(covid2 %in% c('covid-unly', 'covid-othr'), decd_age_yr<=50) %>% group_by(yr) %>%  summarise(count=n())


```

# COVID deaths under age 65
```{r}
deaths %>% filter(covid2 %in% c('covid-unly', 'covid-othr'), decd_age_yr<65) %>%  group_by(yr) %>% summarise(under65 = n())
```


# COVID deaths age 65 and older
```{r}
deaths %>% filter(covid2 %in% c('covid-unly', 'covid-othr'), decd_age_yr>=65) %>%  group_by(yr) %>% summarise(x65plus = n())
```

# COVID deaths age 85 and older
```{r}
deaths %>% filter(covid2 %in% c('covid-unly', 'covid-othr'), decd_age_yr>=85) %>%  group_by(yr) %>% summarise(x85plus = n())
```

